---
title: "Life Satisfaction in South America"
subtitle: "COMP/STAT 112 Final Project Blog"
author: "Maya Harrison, Flor Perez Nunez, Inbal Armony"
date: "2023-12-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(ggplot2)
#library(gplots)
#library(maps) 
library(dplyr)  
#library(ggmap)
library(lubridate)
library(ggthemes)
library(RColorBrewer) 
library(sf)

```

```{r}
load('LAPOP.rda')
LAPOP <- da36562.0001
rm(da36562.0001)

sf_use_s2(FALSE)
#SouthAmericasf <- read_sf("vc965bq8111") %>%
#  group_by(name) %>%
#  summarise(geometry = st_combine(geometry)) %>%
#  mutate(name = stringr::str_to_lower(name))

library(rnaturalearth)
SouthAmericasf <- ne_countries(continent = 'South America',returnclass = 'sf') %>%
  mutate(name = stringr::str_to_lower(admin))

```
```{r}
SouthAmerica <- LAPOP %>% 
  filter(PAIS %in% c("(17) Argentina", "(10) Bolivia", "(08) Colombia", "(15) Brazil", "(12) Paraguay", "(14) Uruguay", "(11) Peru", "(13) Chile", "(09) Ecuador", "(16) Venezuela", "(27) Suriname")) %>% 
  mutate(PAIS = substr(PAIS, 6, 20))
rm(LAPOP)
```

```{r}

# Creating our datasets

LS3_over_time <- SouthAmerica %>% 
  select(YEAR, PAIS, LS3) %>% 
  drop_na() %>% 
  group_by(PAIS, YEAR) %>%
  mutate(LS3num = fct_recode(LS3, "1" = "(1) Very Satisfied", "2" = "(2) Somewhat Satisfied", "3" = "(3) Somewhat Dissatisfied", "4" = "(4) Very Dissatisfied")) %>%
  mutate(total_ls3 = n()) %>% 
  filter(LS3num == "1") %>% 
  mutate(total_good_ls3 = n()) %>% 
  distinct(PAIS, YEAR, .keep_all = TRUE) %>% 
  mutate(percent_good_ls3 = 100 * total_good_ls3 / total_ls3) %>% 
  select(PAIS, YEAR, percent_good_ls3, total_good_ls3, total_ls3)

# Gov protects basic rights
B3_over_time <- SouthAmerica %>% 
  select(YEAR, PAIS, B3) %>% 
  drop_na() %>% 
  group_by(PAIS, YEAR) %>%
  mutate(total_b3 = n()) %>% 
  filter(B3 >= 5) %>% 
  mutate(total_good_b3 = n()) %>% 
  distinct(PAIS, YEAR, .keep_all = TRUE) %>% 
  mutate(percent_good_b3 = 100 * total_good_b3 / total_b3) %>% 
  select(PAIS, YEAR, percent_good_b3, total_good_b3, total_b3)

# Democracy 
DEM2_over_time <- SouthAmerica %>% 
  select(YEAR, PAIS, DEM2) %>% 
  drop_na() %>% 
  group_by(PAIS, YEAR) %>% 
  mutate(total_dem2 = n()) %>% 
  filter(DEM2 == "(2) Democracy is preferable to any other form of government") %>% 
  mutate(total_good_dem2 = n()) %>% 
  distinct(PAIS, YEAR, .keep_all = TRUE) %>% 
  mutate(percent_good_dem2 = 100 * total_good_dem2 / total_dem2) %>% 
  select(PAIS, YEAR, percent_good_dem2, total_good_dem2, total_dem2)

#Economic situation 
SOCT1_over_time <- SouthAmerica %>% 
  select(YEAR, PAIS, SOCT1) %>% 
  drop_na() %>% 
  group_by(PAIS, YEAR) %>% 
  mutate(total_soct1 = n()) %>% 
  filter(SOCT1 %in% c("(2) Good", "(1) Very Good")) %>% 
  mutate(total_good_soct1 = n()) %>% 
  distinct(PAIS, YEAR, .keep_all = TRUE) %>% 
  mutate(percent_good_soct1 = 100 * total_good_soct1 / total_soct1) %>% 
  select(PAIS, YEAR, percent_good_soct1, total_good_soct1, total_soct1)

questions_over_time <- left_join(LS3_over_time, B3_over_time)
questions_over_time <- left_join(questions_over_time, DEM2_over_time)
questions_over_time <- left_join(questions_over_time, SOCT1_over_time)
questions_over_time <- questions_over_time %>% 
  mutate(YearCat = cut(YEAR, c(2004,2009,2015), right = FALSE, labels = c('2004-08','2009-14')))

```

When you think about your life, what influences your life satisfaction the most? Is it your friends and family, your job, or maybe even economic stability? Would the people around you answer the same thing? 

In a joint research project called the Latin American Public Opinion Project (LAPOP), USAID, and Vanderbilt University surveyed citizens of 28 Latin American countries using a national probability design. In one question the researchers asked participants, “In general, how satisfied are you with your life?” Over 10 years (2004-2014) voting-age adults responded to this question, along with questions about their impressions of their nation’s economic situation, democracy, basic rights, and many other topics. 

Looking through the data, we narrowed down the data set to the eleven surveyed countries in South America. After several conversations comparing our experiences in the United States with life experiences in South America, our group was curious to find out what determines life satisfaction in South American countries. And more specifically, between perceptions of basic rights, democracy, and the economic situation, which most correlates to life satisfaction? For these questions, each country had between 1,100 and 3,000 respondents per year.

#### Life satisfaction differs throughout South America

```{r}
# LIFE SATISFACTION map animination (In general how satisfied are you with your life?) (1 = very satisfied, 4 = very dissatisfied)

library(gganimate)
SouthAmericaLS3 <- SouthAmerica %>% 
  mutate(PAIS = stringr::str_to_lower(PAIS)) %>%
  select(YEAR, PAIS, LS3) %>% 
  drop_na() %>% 
  mutate(LS3num = fct_recode(LS3, "1" = "(1) Very Satisfied", "2" = "(2) Somewhat Satisfied", "3" = "(3) Somewhat Dissatisfied", "4" = "(4) Very Dissatisfied")) %>%
  count(PAIS,YEAR,GoodLS3 = LS3num == "1") %>%
  group_by(PAIS,YEAR) %>%
  mutate(percent_good_ls3 = 100 * n / sum(n))  %>%
  filter(GoodLS3) 


questions_over_time <- questions_over_time %>% 
  mutate(name = stringr::str_to_lower(PAIS))
  
questions_over_time_MAP <- questions_over_time %>%
right_join(expand_grid(name =unique(SouthAmericasf$name), YEAR = unique(questions_over_time$YEAR)) %>% filter(!is.na(YEAR))) %>% 
select(YEAR, name, percent_good_ls3)

SouthAmericas_map_data <- SouthAmericasf %>% left_join(questions_over_time_MAP, by = c('name' = 'name'))

#These two lines, run in the console, fix the error about crs
#install.packages('devtools')
#devtools::install_version("transformr", version = "0.1.3")
library(transformr)
library(gganimate)
library(gifski)
map_anim <- 
  SouthAmericas_map_data %>% 
  arrange(YEAR) %>%
  mutate(LS3_Cat = factor(percent_good_ls3)) %>% 
  mutate(LS3_Cat = fct_rev(LS3_Cat)) %>% 
  mutate(LS3_Cat = fct_na_value_to_level(cut(percent_good_ls3,5),level='No Data')) %>%
  ggplot() +
  geom_sf(aes(fill = LS3_Cat),linewidth = .05) +
  labs(fill = '% of Respondents with High Life Satisfaction') +
  scale_fill_manual(values = c(colorRampPalette(c("white","darkgreen"))(5),'lightgrey')) +
  theme_map() +
  theme(legend.background = element_blank()) +
  theme(legend.position = "right") +
  transition_states(YEAR) + labs(title = 'High Life Satisfaction in South America: {closest_state}')

animate(map_anim, nframes = 20, fps = 1, renderer = gifski_renderer())

anim_save("ls3_map.gif",map_anim)

```

We wanted to begin by geographically contextualizing each country's life satisfaction. This map of South America displays the percentage of respondents who reported having a very high life satisfaction in each country, over time. Countries shaded with the darkest green, like Brazil in 2008, had between 52-63% of respondents report a very high life satisfaction, while countries shaded with the lightest green, like Bolivia in 2007, had only 20-31% of respondents report very high life satisfaction. Countries with gray shading had no data for that year. Some years have many countries not surveyed, highlighting the sparsity of our data. Critically, the map shows just how much variation there is in life satisfaction between countries and over time in South America.

The map provides important geographic information that could impact life satisfaction, like coastal versus land-locked countries and possible relationships between neighboring countries. This map portrays some interesting geographic trends in this data like the Northern and warmest countries of the continent (Brazil, Colombia, and Venezuela) consistently having higher percentages of high life satisfaction over time. In contrast, between 2008-2012, southern countries like Bolivia, Chile, and Peru had much fewer respondents with high life satisfaction. 

#### Every South American country experiences fluctuations in life satisfaction over time

```{r}
# LS3 all countries over time animation - Inbal
library(gganimate)
pp_anim <- questions_over_time %>% 
  select(YEAR, PAIS, percent_good_ls3, percent_good_b3, percent_good_dem2, percent_good_soct1) %>% 
  pivot_longer(3:6, names_to = "Question", values_to = "Percent Good", values_drop_na = FALSE) %>% 
  filter(Question == "percent_good_ls3") %>% 
  ggplot(aes(x=YEAR, y=`Percent Good`, color=PAIS, group = PAIS)) +
  geom_point() + geom_line() + theme_classic() +
transition_reveal(YEAR) + labs(title = "High Life Satisfaction Over Time in South America", y="Percent of Respondents who Reported High Life Satisfaction", x="Year", color="Country")

animate(pp_anim, renderer = gifski_renderer())

anim_save("ls3_scatter_plot.gif", pp_anim)

```

Our next graph displays the same data as the previous map––percentages of respondents reporting high life satisfaction––but shown on a scatter plot that is animated over time. Each individual country is represented by its own, different colored line and the animation helps highlight change over time of life satisfaction. This visualization helps with comparisons between nations as we can easily see when one nation has a greater portion of respondents with high life satisfaction than another nation. 

Venezuela, Brazil, and Colombia all maintain high levels of respondents with high life satisfaction above 40-50%. Interestingly, these three countries and some others all experienced a decrease in life satisfaction rates in 2012. At the bottom of the graph, Bolivia and Peru only have 20-30% of respondents with high life satisfaction. The purple line of Suriname stands out for having a great increase in life satisfaction in the year 2012, but this might be attributed to low survey size and data from only two years of the survey. 

#### Bolivia, Uruguay, Peru, and Brazil with Most Consistent Life Satisfaction 

```{r}
# Scatter plot, faceted by country, LS3 over time - Maya

questions_over_time %>%
  ggplot(aes(x= YEAR, y=percent_good_ls3)) +
  geom_point(color = "darkmagenta") +
  geom_line(color = "darkmagenta") +
  facet_wrap(~PAIS) +
  labs(x = "Year", y = "Respondents who Reported High Life Satisfaction (%)", title = "Positive Perception of Life Satisfaction Over Time in South America")+
   theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust=.7))
```

While many countries demonstrated great fluctuation in perceptions of life satisfaction, others showed consistent levels over time. In this graph, we can see that Argentina, Chile, Ecuador, Paraguay, and Suriname had great variation in their responses over time. By contrast, Bolivia, Brazil, Colombia, Ecuador, Uruguay, and Venezuela had consistent percentages of respondents reporting high life satisfaction. Venezuela, Brazil, and Colombia stand out for having the most consistent high life satisfaction, while Bolivia had the most constant low life satisfaction. This made us question why Bolivia had such a consistently low life satisfaction and whether other perceptions, particularly of the economy, protection of basic rights, and democracy contribute to low life satisfaction in the country. 


#### Perceptions of life satisfaction and the economy change together

```{r}
# flor 

questions_over_time_Brasil_soct1 <- tibble( x = 34.636119, y = 55.07343)
questions_over_time_Peru_soct1 <- tibble( x = 18.218352, y = 31.55765)
questions_over_time_Chile_soct1 <- tibble( x = 26.228462, y = 42.48241)
questions_over_time_Paraguay_soct1 <- tibble( x = 19.148936, y = 48.79679)

library(gganimate)
pp_anim <- questions_over_time %>%
  filter(PAIS == "Bolivia") %>%
  ggplot() +
    geom_point(aes(y=percent_good_ls3, x=percent_good_soct1, group=PAIS)) + geom_path(aes(y=percent_good_ls3, x=percent_good_soct1, group=PAIS),color = "darkmagenta") +
  geom_text(data = questions_over_time_Brasil_soct1, aes(y=y, x=x, label = "Brasil")) +
  geom_text(data = questions_over_time_Peru_soct1, aes(y=y, x=x, label = "Peru")) +
  geom_text(data = questions_over_time_Chile_soct1, aes(y=y, x=x, label = "Chile")) +
  geom_text(data = questions_over_time_Paraguay_soct1, aes(y=y, x=x, label = "Paraguay")) +
  theme_classic() +
  labs(title = "Bolivia's High Life Satisfaction and Economy", subtitle = "From 2004-2014", x = "Percentage of  Participants who Reported Good/Very Good Economic Situation" ,y = "Percentage of Participants who Reported High Life Satisfaction") +
 transition_reveal(YEAR)

animate(pp_anim, renderer = gifski_renderer())

anim_save("ls3_soct1", pp_anim)
```


Upon looking closely at Bolivia's percentage of high life satisfaction and the perception of a favorable economic situation, a pattern emerges. Initially, both indicators experienced an upward trajectory, signaling positive trends. However, in the subsequent period, there was a noticeable decline in both measures. The perception of the economic situation improved, with an increase of approximately 25 percentage points. This suggests a positive shift in the way individuals in Bolivia perceive their economic circumstances. Simultaneously, life satisfaction showed an increase of around 10 percentage points, indicating an overall positive sentiment among the populace. When juxtaposed with neighboring countries, Bolivia's standings in both categories are comparatively lower. This implies that, despite the observed improvements, Bolivia stays behind its regional counterparts in terms of both economic optimism and overall life satisfaction. 
	
#### Democracy Inversely Correlated to Life Satisfaction 

```{r}
# Scatter plot of Bolivia with all 4 questions over time - Flor
questions_over_time_Bolivia <- questions_over_time
colnames(questions_over_time_Bolivia)[colnames(questions_over_time_Bolivia) == "percent_good_ls3"] <- "High life satisfaction"
colnames(questions_over_time_Bolivia)[colnames(questions_over_time_Bolivia) == "percent_good_b3"] <- "Govt. protects basic rights"
colnames(questions_over_time_Bolivia)[colnames(questions_over_time_Bolivia) == "percent_good_dem2"] <- "Democracy is most preferable"
colnames(questions_over_time_Bolivia)[colnames(questions_over_time_Bolivia) == "percent_good_soct1"] <- "Good economic situation"

questions_over_time_Bolivia %>%
  select(c("YEAR", "PAIS", "Good economic situation", "Govt. protects basic rights", "Democracy is most preferable", "High life satisfaction")) %>%
  pivot_longer(3:6, names_to = "Categories", values_to = "Percent Good", values_drop_na = FALSE) %>%
  filter(PAIS == "Bolivia") %>%
  ggplot(aes(x=YEAR, y=`Percent Good`, color=Categories)) +
  geom_point() + geom_line() + 
  labs(title = "Bolivia's Positive Perceptions", subtitle = "Life Satisfaction, Democracy, Basic Rights, and the Economy From 2004 - 2014", x= "Year", y= "Pecrentage of Agreement",fill = '') +
  theme_classic() + scale_color_brewer(palette = "pastel2") +
  scale_color_manual(values = c("Good economic situation" = "darkorchid4", "Govt. protects basic rights" = "chocolate4", "Democracy is most preferable" = "chocolate1", "High life satisfaction" = "darkturquoise" )) 
```

This graph illustrates changes over time for four distinct questions derived from an opinion survey. It delineates how Bolivian respondents expressed their views on having a good economic situation in their country, their preference for democracy as the form of government, the percentage of respondents who believed their government protected basic rights, and their level of life satisfaction. Overall, there appears to be a straightforward, inverse relationship between democracy and the other three preferences. For instance, between 2004 and 2006, the percentage of respondents indicating democracy as the most preferable form of government fell, while the percentage of positive responses to the other questions rose.

This observation gains significance when considering the Bolivian presidential election in 2005, resulting in the election of a new president, Evo Morales. Between 2010 and 2012, there was a decline in life satisfaction, perceptions of a good economic situation, and confidence in the government’s protection of basic rights. Interestingly, the perception of democracy as the preferred form of government remained stable. The decrease in these three perceptions aligns with the political and economic instability during that period. In 2008, an economic crisis hit Bolivia and other South American countries. Simultaneously, there was political instability marked by protests against President Morales, escalating into the Porvenir massacre as a consequence of violent right-wing protests.

#### Conclusion

Overall, between democracy, basic rights, and the economy, no single perception determines life satisfaction in South America. However, exploring individual countries in depth, as we did with Bolivia, reveals interesting relationships between the perceptions.

It is critical to note that these data points represent the lived experiences of real people. Complex experiences cannot be accurately quantified or generalized for an entire country, making this dataset inherently flawed. The conclusions we drew about life satisfaction in Latin America provide important starting off points for more research but cannot be taken as conclusive evidence. Rather, we hope to honor these real experiences by presenting the complexities of the question of life satisfaction.

Bolivia’s lower rates of high life satisfaction, by comparison with other nations in South America, must be understood within the country’s complex political and economic landscape. The similar changes in perceptions of life satisfaction, basic rights, and the economy provide insight into the impact of intense moments of political strife and economic crisis. More specific perceptions and factors––such as Bolivia being a landlocked country, or ratings of Morales over time––should be explored in future research to better understand the issues more impacting Bolivians. 

#### References:

Congressional Research Service. (n. d.) https://crsreports.congress.gov/product/pdf/IF/IF11325

Center for Economic and Policy Research. (2019). Bolivia’s Macroeconomic Situation.   https://cepr.net/images/stories/reports/bolivia-macro-2019-10.pdf

Thomson, S., Barragán, R., Albó, X., Qayum, S., & Goodale, M. (Eds.). (2018). The Bolivia Reader: History, Culture, Politics. Duke University Press. https://doi.org/10.2307/j.ctv11313s0

Vanderbilt University. (n.d.). Latin American Public Opinion Project (LAPOP) Survey Data. Retrieved from https://www.vanderbilt.edu/lapop/survey-data.php


